---
title: 'Module 1.1: Importing and Exporting Vector GIS Data in R'
author: "IALE 2015 R GIS Student Workshop"
date: "July 07, 2015"
output:
  html_document:
    toc: yes
    keep_md: true
---

Module 1.1 code by Jillian Deines, Michigan State University

```{r knitrOptions, include=FALSE}
library(knitr)
opts_chunk$set(cache=TRUE, fig.path='figures/m1.1/')
opts_knit$set(root.dir = 'D:/RWorkshop/Intro-Spatial-R/module1')
```

## R Packages Needed
__Workhorse packages:__ You can do a whole heck of a lot with a few basic packages:

```{r loadPackages, message=FALSE, warning=FALSE}
library(sp)         # a dependency for most spatial packages (also loaded with rgdal)
library(rgdal)      # R bindings for the Geospatial Data Abstraction Library    
library(maptools)   # for the 'readShapeSpatial' example
```

## Set Your Working Directory
Here, we'll use the `setwd` function to tell R where the module 1 folder is located on our machine. This will depend on the location of your downloaded course folder, Intro-Spatial-R.

```{r setwd}
# modify the filepath based on your user specific folder location
# Single or double quotes work for text strings. Must use forward slashes.

#Some common examples:
#setwd('C:/Users/username/Documents/Intro-Spatial-R/module1')   # for Windows users
#setwd("~/Documents/Intro-Spatial-R/module1")                   # ~ for Mac users
setwd('D:/RWorkshop/Intro-Spatial-R/module1')                   # Jill's computer  
```


## Read in Shapefiles
Two primary functions exist to read in .shp files:

* `readShapeSpatial` in package 'maptools' doesn't read projection information; need to set manually
* `readOGR` in package 'rgdal' reads projection information, so **we'll use this exclusively**. More information on manipulating projections in R will follow in Module 2.

### Import Polygon Shapefile
Here, we import a polygon boundary for the Colorado Plateau (CP) that has an Albers Equal Area projection, "COP_boundpoly_aea.shp" from the 'module1/data' course subfolder. 

```{r loadSHP2}
# load shapefile by specifying folder name (dsn), and layer name (no extension!)
cp.poly <- readOGR(dsn = 'data',             # subfolder name 
                   layer = 'COP_boundpoly_aea')   # file name without extension
```

#### Examples for Alternative Directory Structures
Many people do not use `setwd` to set working directories, or you may have your directory structured differently on future projects. Here are a few examples for alternative directory structures. The examples are specific to my directory structure, so these examples are not run.

**1. Not using a working directory**  
This is essentially what `setwd` does under the hood - appends the working directory filepath to your directory call:

```{r loadSHP3, eval=FALSE}
cp.poly <- readOGR(dsn = 'D:/RWorkshop/Intro-Spatial-R/module1/data',
                   layer = 'COP_boundpoly_aea')
```

**2. Shapefile located in your base working directory**  
If you are reading in a shapefile that is located in your working directory, you need to use `dsn = '.'` to refer to the current working directory:

```{r loadSHP4, eval=FALSE}
setwd('D:/RWorkshop/Intro-Spatial-R/module1/data') 
cp.poly <- readOGR(dsn = '.', layer = 'COP_boundpoly_aea')
```

#### Examine loaded shapefile
Let's take a look at what we loaded.

```{r polyExplor}
# View description. Note the coordinate bounding box and projection info
summary(cp.poly)
# just look at projection
proj4string(cp.poly)
# make a simple plot
plot(cp.poly, axes=T, col='blue', main = "CO Plateau")
```

#### Alternative demonstration: readShapeSpatial in package Maptools

If you search for help online, you may find another common way to load shapefiles: the `readShapeSpatial` function in the package maptools. This is a valid alternative, but it's important to note that this function does not automatically load projection information:

```{r maptoolsExample}
# load shapefile by specifying file name with extension
cp.poly2 <- readShapeSpatial('data/COP_boundpoly_aea.shp') 

## NOT RUN: equivalent alternative example if not setting a working directory,
## based on my computer's directory structure:
#cp.poly2 <- readShapeSpatial('D:/RWorkshop/Intro-Spatial-R/module1/data/COP_boundpoly_aea.shp')

# look at the projection information: none
summary(cp.poly2)
```

Instead, you need to know your projection and define it manually. This does not involve re-projecting your GIS layer, or changing the projection. You are only telling R what projection your file is in. R uses PROJ.4 projection definitions, which you can find by searching at the website http://spatialreference.org. More on this in module 2. Here we'll assume I know the proj4 information. **Note that proj4strings are sensitive to spacing and can have no hard returns/line breaks**.

```{r maptoolsEx2}
# define the AEA projection information
aea.proj <- "+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=23 +lon_0=-96 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs +ellps=GRS80 +towgs84=0,0,0"
# set projection of object
proj4string(cp.poly2) <- CRS(aea.proj)
# look at projection info
summary(cp.poly2)
```

I prefer the single step of `readOGR`, but this is helpful to know if the shapefile itself doesn't have a defined projection and you need to define it manually.

### Import Line and Point Shapefiles
All shapefiles are read in the same with the `readOGR` function (as well as `readShapeSpatial`). Here we'll load points from the "COP_Major_Cities_pt.shp" as a demonstration. Note that this shapefile has an attribute table with 47 columns/attributes.

```{r loadLinesPoints}
# load shapefile by specifying folder name (dsn), and layer name (no extension!)
cities <- readOGR(dsn = 'data', layer = 'COP_Major_Cities_pt')
# make a quick plot. note that projections need to match
par(mar=c(2,2,2,2))             # make plot margins smaller
plot(cp.poly, col='azure2')     # plot CO Plateau polygon
plot(cities, add=T,             # add city points using add=T
     col='red',                 # point color      
     pch=19,                    # point type (here, solid circles)
     cex=1.5)                   # point size
```

## Convert Latitude/Longitude Spreadsheet Data to A Point Shapefile
If you have a spreadsheet of latitude and longitude data points, you can convert that to a spatialPointsDataFrame (spdf) in R and then write out a shapefile for future use. Here we'll use coordinate data that is in .csv format, but you could also use .xls(x) provided you use the appropriate R package to load Excel files (I'd recommend `readxl`). Generally, .csv or .txt files are faster and easier to use.

Here, we'll load lat/long presence/absence data for the species Juniperus osteosperma in the juos_presab.csv spreadsheet and learn how to write out shapefiles.

```{r convertXY, echo=-1} 
# load xy data as a data frame
juos <- read.csv('data/juos_presab.csv')
# look at data structure: 
str(juos)
head(juos)
# convert dataframe to spdf by specifying long and lat column names
# format: ~ longitudeColumn + latitudeColumn
coordinates(juos) <- ~ wgs84_x + wgs84_y
# check out our new spdf object
summary(juos)
```

Note once again that our projection is undefined. Since the .csv has no associated projection information, we need to specify it. We know that the coordinates are in a WGS84 based lat/long coordinate reference system. Again you can use http://spatialreference.org to find proj4 strings (more detail in Module 2)

```{r convertXY2}
# create character vector defining the proj4 string for WGS84
WGS84.proj <- "+proj=longlat +datum=WGS84" 
# set the proj4string
proj4string(juos) <- CRS(WGS84.proj)  
# take a look
summary(juos)
```

Note that 'Is projected:" is FALSE because we are using latitude/longitude coordinates, not a projected reference system based on meters, etc.

Plot our new spatial points, coloring based on presence/absence attribute.

```{r plotxy}
# make a plot based on juos attribute column (0 or 1)
plot(juos, pch=19, cex = 0.8, col=c("cadetblue3","coral1"))
# add a simple legend
legend("topleft",                          # position in plot
       fill=c("cadetblue3","coral1"),      # match colors with plot colors   
       legend = c("0","1"),                # specify legend text
       bty='n')                            # don't use a box around the legend
```

## Write Out Shapefile
All shapefiles (points, lines, and polygons) can be written out with the `writeOGR` function in the rgdal package. Here we demonstrate with our newly created spatial point object, which we will write out to the "outData" subfolder within the module 1 data directory and call "juos_pts_wgs84.shp".

Note that we will use the ESRI shapefile driver to create a .shp file, but the 'driver' argument has many options including GeoJSON, KML, etc. Typing ogrDrivers() at your command line will give a list of available drivers.

```{r writePoints}
# write out point shapefile. On Windows at least, new subfolder can be created in call
writeOGR(obj = juos, dsn = 'data/outData', layer = 'juos_pts_wgs84',
         driver = 'ESRI Shapefile', overwrite_layer=T)
```

There should now be a file "juos_pts_wgs84.shp" in your ...IALE2015_RCOURSE\data\outdata folder. 

End Module 1.1.

```{r eval=FALSE, include=FALSE}

rmd2rscript <- function(infile){
  # read the file
  flIn <- readLines(infile)
  # identify the start of code blocks
  cdStrt <- which(grepl(flIn, pattern = "```{r*", perl = TRUE))
  # identify the end of code blocks
  cdEnd <- sapply(cdStrt, function(x){
    preidx <- which(grepl(flIn[-(1:x)], pattern = "```", perl = TRUE))[1]
    return(preidx + x)
  })
  # define an expansion function
  # strip code block indacators
  flIn[c(cdStrt, cdEnd)] <- ""
  expFun <- function(strt, End){
    strt <- strt+1
    End <- End-1
    return(strt:End)
  }
  idx <- unlist(mapply(FUN = expFun, strt = cdStrt, End = cdEnd, 
                SIMPLIFY = FALSE))
  # add comments to all lines except code blocks
  comIdx <- 1:length(flIn)
  comIdx <- comIdx[-idx]
  for(i in comIdx){
    flIn[i] <- paste("# ", flIn[i], sep = "")
  }
  # create an output file
  nm <- strsplit(infile, split = "\\.")[[1]][1]
  flOut <- file(paste(nm, "[rmd2r].R", sep = ""), "w")
  for(i in 1:length(flIn)){
    cat(flIn[i], "\n", file = flOut, sep = "\t")
  }
  close(flOut)
}

infile <- 'C:/Users/deinesji/Dropbox/1PhdJill/miscPresentations/IALE_2015/RWorkshop/WorkshopCode/M1.1-ImportExport-Shapefiles.Rmd'

rmd2rscript(infile)
```
