# Module 5: Interactive online mapping and visualization using R

**author: Francesco Tonini**  
**email: ftonini84@gmail.com**  
**date: June 8, 2015**  
**event: IALE 2015 R GIS Workshop**

### Goal: Learn how to create interactive online maps with R

## R Packages Needed

```{r load packages, message = FALSE}
library(rgdal)   #for reading/writing geo files
library(rgeos)   #for simplification
library(sp)
library(plyr)
library(dplyr)
library(leafletR)
library(maptools)
library(RColorBrewer)
library(Quandl)
library(reshape2)
```

```{r load devtools, message = FALSE}
library(devtools)
##The R packages rCharts and rMaps are not available on CRAN yet.
##Run the following code to install from Github repo prior to loading the libraries:
##install_github('ramnathv/rCharts')
##install_github('ramnathv/rMaps')
library(rCharts)
library(rMaps)
```

## Set Your Working Directory

Change this to the correct path name on your machine

```{r set working directory, eval=FALSE}
setwd("D:/Google Drive/IALE2015/r_code/module5")
#setwd("path_to_your_folder")   # for Windows users
#setwd("~/path_to_your_folder"") # ~ for Mac users
```

## *Module 5.1: Create an interactive online map using the leafletR package*

## Read shapefiles

Here we import a polygon boundary of the state of Utah and a set of polygons representing game management units within the state.

```{r read shapefiles}
deer_unit <- readOGR(dsn = "data", layer = "deerrange_UT") # Game management units
```

The attribute data is stored in a slot called `@data` so if we just want to look at the data stored in the object:

```{r summary shapefile data}
summary(deer_unit@data)
names(deer_unit) # The attribute data names
class(deer_unit)
```

## Calculate a new field for vector data

Calculating a new field is as easy as adding a new column to any other data frame in `R`. Say we want to calcualte a new field from the existing data, such as the total number of adult deer and elk in each management unit:

```{r calculate new field}
deer_unit$tot_deer_elk <- rowSums(cbind(deer_unit$DEER, deer_unit$ELK))
summary(deer_unit@data)
```

## Simplify your shapefile (if necessary)

Many shapefiles have significant detail that results in very large GeoJSON files. We will use the rgeos library to simplify. Note: if preserving topology is crucial, consider using tools in QGIS or GRASS GIS. For this demonstration we will restrict ourselves to R

```{r simplify shapefile}
#save the data slot
deer_unit_sub <- deer_unit@data[,c("UName", "UNum", "AREA_km2", "tot_deer_elk")]
#simplification yields a SpatialPolygons class
deer_unit <- gSimplify(deer_unit, tol=0.01, topologyPreserve=TRUE)
class(deer_unit)
#to write to geojson we need a SpatialPolygonsDataFrame
deer_unit <- SpatialPolygonsDataFrame(deer_unit, data=deer_unit_sub)
class(deer_unit)
head(deer_unit@data)
```

## Use leafletR to produce the final online map

We are now ready to create the GeoJSON file. We will create the cuts we want to map and then create the Leaflet map

```{r write GeoJSON}
#write data to GeoJSON
dir <- paste(getwd(), "DeerElkGeoJson", sep="/")
writeOGR(deer_unit, dir, layer="DeerElk", driver="GeoJSON")
#a GeoJSON datasource is translated to single OGRLayer object with pre-defined name OGRGeoJSON"
ogrListLayers(paste(getwd(), "DeerElkGeoJson", sep="/"))
ogrInfo(paste(getwd(), "DeerElkGeoJson", sep="/"), "OGRGeoJSON")
```

Create the cuts

```{r create cuts}
cuts <- round(quantile(deer_unit$tot_deer_elk, probs = seq(0, 1, 0.20), na.rm = FALSE), 0)
```

Choose fields to include in the popup

```{r popup fields}
popup <- c("UName", "UNum", "AREA_km2", "tot_deer_elk")
```

Graduated style symbology based on an attribute

```{r graduated style}
sty <- styleGrad(prop="tot_deer_elk", breaks=cuts, closure="left", style.par="col", 
                 style.val=rev(heat.colors(5)), leg="Deer & Elk Population", lwd=1)
```

Create the map and load into browser

```{r create map, message=FALSE}
map <- leaflet(data=dir, dest=getwd(), style=sty,
             title="DeerElk", base.map="osm",
             incl.data=TRUE,  popup=popup)
```

To look at the map you can use this code

```{r browse map}
browseURL(map)
```

## *Module 5.2: Create Leaflet heat maps*

In this module, we will work on a dataset containing abundance records of bay laurel (UMCA) trees in the Sonoma county, California. After reading in and cleaning the dataset, we will create a leaflet heat map using a JavaScript plugin.

## Get and subset data
```{r get data}
#read in data table 
tree_dat <- read.csv("./data/plots_umca_infection.csv", header=T)
#read plot locations shapefile
plots <- readOGR(dsn='./data', layer='plot_202_latlon')
#using the `dplyr` package by Hadley Wickham, let's subset data for year 2012 only.
tree_dat_2012 <- tree_dat %>%
     select(plot, year, tot_bay) %>%
     group_by(plot) %>%
     filter(year == 2012)
tree_dat_2012
```

## Add geographic coordinates

We have 202 plot locations in the shapefile but our dataset only has 179. Let's subset coordinates only for the matching ones.

```{r add latlon}
name_match <- match(tree_dat_2012$plot, plots@data$PLOT_ID)
tree_dat_2012 <- cbind(plots@data[name_match, c("POINT_Y","POINT_X")], tree_dat_2012)
names(tree_dat_2012)[1:2] <- c("lat", "lon") 
#get rid of the year and plot variables.
tree_dat_2012 <- tree_dat_2012[ ,!names(tree_dat_2012) %in% c("year","plot")]
head(tree_dat_2012)
```

## Create new Leaflet map

Create new Leaflet map and plot study area centered on our coordinates

```{r leaflet_map, message=FALSE}
#create a new leaflet map instance
Lmap <- Leaflet$new()
#set the view and zoom to the desired study area. Let's center it on our mean lat-lon coordinates
Lmap$setView(c(mean(tree_dat_2012$lat), mean(tree_dat_2012$lon)), 10)
#add a basemap using OSM 
Lmap$tileLayer(provider = "MapQuestOpen.OSM")
#plot the study area (.html file is created locally)
Lmap
```

## Convert data to JSON format
```{r convert to JSON}
tree_dat <- toJSONArray2(na.omit(tree_dat_2012), json = F, names = F)
#let's print out the first two elements of the JSON file
cat(rjson::toJSON(tree_dat[1:2]), '\n')
```

## Add leaflet-heat plugin and create heat maps
```{r heat_map, message=FALSE}
#add leaflet-heat plugin. Thanks to Vladimir Agafonkin
Lmap$addAssets(jshead = c(
  "http://leaflet.github.io/Leaflet.heat/dist/leaflet-heat.js"
))

#add javascript to modify underlying chart
Lmap$setTemplate(afterScript = sprintf("
<script>
  var addressPoints = %s
  var heat = L.heatLayer(addressPoints).addTo(map)           
</script>
", rjson::toJSON(tree_dat)
))
#plot heat map of UMCA tree abundance (.html file is created locally)
Lmap
```

## *Module 5.3: Create animated choropleths*

The following tutorial is based on the [rMaps] (http://rmaps.github.io/blog/posts/animated-choropleths/index.html) online tutorial used to create animated choropleths. The data selected are different from the original. 

## Get Data

The first step to creating any visualization is getting the data. Let us fetch time-series data on burglaries in the US, from [Quandl] (https://www.quandl.com/). 

```{r get Quandl data}
rbData = Quandl("FBI_UCR/USCRIME_TYPE_BURGLARIES")
rbData[1:10, 1:5]
```

## Reshape the data into wide-form

The dataset is in the wide-form. The first step is to convert it into the long-form (more convenient for visualization purposes). Moreover, we remove data for the US as a whole, as well as for DC, so that the crime rates across entities (states) are comparable

```{r reshape data}
datm <- melt(rbData, 'Year', 
  variable.name = 'State',
  value.name = 'Crime'
)
datm <- subset(na.omit(datm), 
  !(State %in% c("United States", "District of Columbia"))
)
head(datm)
```

## Categorize crime rates

Crime rates need to be discretized. One way to do this is to divide them into quartiles

```{r discretize data}
datm2 <- transform(datm,
  State = state.abb[match(as.character(State), state.name)],
  fillKey = cut(Crime, quantile(Crime, seq(0, 1, 1/4)), labels = LETTERS[1:4]),
  Year = as.numeric(substr(Year, 1, 4))
)
```

## Associate classification with fill colors

Each quartile needs to be associated with a fill color chosen from a palette

```{r fill color}
fills = setNames(
  c(RColorBrewer::brewer.pal(4, 'OrRd'), 'white'),
  c(LETTERS[1:4], 'defaultFill')
)
```

## Convert to data structure for the `DataMaps` library

We need to convert the data frame into a list of lists. We will use Hadley's `plyr` package to simplify the code

```{r convert data frame}
dat2 <- dlply(na.omit(datm2), "Year", function(x){
  y = toJSONArray(x, json = F)
  names(y) = lapply(y, '[[', 'State')
  return(y)
})
names(dat2)
#dat2[["1960"]] to inspect a list element
```

## Create a simple choropeth map

We can now create a simple choropleth map of crime rates for a given year. The Datamaps reference class gives us simple bindings to the DataMaps library

```{r simple choro map}
options(rcharts.cdn = TRUE)
map <- Datamaps$new()
map$set(
  dom = 'chart_1',
  scope = 'usa',
  fills = fills,
  data = dat2[["1980"]],
  legend = TRUE,
  labels = TRUE
)
map
```

## Create a dynamic choropeth map

Use a customized wrapper function that absorbs js code to produce a dynamic choropeth map

```{r dynamic choro map, echo=-1}
setwd("D:/Google Drive/IALE2015/r_code/module5")
source('ichoropleth.R')
map2 <- ichoropleth(Crime ~ State,
  data = datm2[,1:3],
  pal = 'OrRd', #color ramp
  ncuts = 4,  #quartiles
  animate = 'Year'
)
map2
```




