---  
title: "Module 4: Extract and Analyze GIS Data"  
author: "IALE 2015 R GIS Workshop"	
date: "July 7, 2015"	
output:
  html_document:
    keep_md: yes
    pandoc_args:
    - +RTS
    - -K64m
    - -RTS
    toc: yes
  pdf_document:
    toc: yes
---	

# Module 4.1
## Goals: Extract data for juniper across raster and vector layers at a set of points
Extract values from rasters and vectors at a set of points, and import point data for plant locations and rasters representing topographic data

## R Packages Needed  
```{r load packages, message = FALSE}
library(sp)	
library(rgdal)	
library(raster)
library(fields)
```	
	
## Set Your Working Directory	
Change this to the correct path name on your machine	
```{r set working directory}
setwd("~/Documents/IALE2015_StudentWorkshop/module4")
```

## Import Data
### Points
Load shapefile of *Juniperus osteosperma* point data that you created in module 1
```{r load juos points}
juos <- readOGR(dsn = "data/outdata", layer = "juos_pts_wgs84")  
```

### Rasters
Load rasters for elevation, slope and aspect
```{r load rasters}
elev1k.wgs <- raster("data/elev_1k_wgs.img")
slp1k.wgs <- raster("data/slp_1k_wgs.img")
asp1k.wgs <- raster("data/asp_1k_wgs.img")
```

Combine rasters into single object. Earlier, you used RasterStacks. This time, let's use a RasterBrick. They are very similar, but RasterBricks can be faster for computation. RasterStacks can access layers without keeping them all in memory, so that can be an advantage, too. It just depends on your needs.
```{r rasterbrick}
topo1k.wgs <- brick(elev1k.wgs, slp1k.wgs, asp1k.wgs)
```

Take a look at the brick.
```{r view rasterbrick}
topo1k.wgs
```

Shorten the names of the layers in the brick.
```{r change names in rasterbrick}
names(topo1k.wgs) <- c("elev", "slp", "asp")
topo1k.wgs
```

### Vectors
Load shapefile for soil
```{r load soil shapefile}
soil <- readOGR(dsn = "data", layer = "soils")
```

Take a look at what kinds of data are in the `soil` object
```{r summary of shapefile}
summary(soil)
```

Load the polygon that bounds the Colorado Plateau. It will be useful for cropping and masking.
```{r load CO PLateau Polygon}
cp.latlong <- readOGR(dsn = "data", layer = "COP_boundpoly_wgs84")
```

## Crop and Mask Data
### Rasters
Crop elevation raster to a bounding box that fits around the Colorado Plateau
```{r crop rasters}
topo.bbox <- crop(topo1k.wgs, cp.latlong)
```

Mask elevation raster to include only data within the Colorado Plateau polygon. This may take a while.
```{r mask rasters}
topo.cp <- mask(topo.bbox, cp.latlong)
```

### Vectors
In the case of vector data, you can still use the `crop` function. However, it does in one function what both `crop` + `mask` do for raster data.
```{r crop vector data}
soil.cp <- crop(soil, cp.latlong)
```

## Visualize Rasters, Vectors and Points
Take a look at the rasters.
```{r plot rasters}
plot(topo.cp)
```

Plot individual raster with points overlaid. Use `add = TRUE` or `add = T` to add points to the same plot.
```{r plot single raster with points}
plot(topo.cp$elev)
plot(juos, cex = 0.3, pch=19 ,col=c("red","blue"), add=T)
legend("topleft",fill=c("red","blue"), legend = c("0","1"), bty="n")
```

This will fill all the polygons in the soil object with orange.
```{r plot soil polygons in orange}
plot(soil.cp, col="orange")
```

Alternatively, we can fill it with colors corresponding to data from one of the soil variables.
```{r plot soil polygons by data}
plot(soil.cp, col=soil.cp$taxorder)
plot(juos, cex = 0.3, pch=19 ,col=c("red","blue"), add=T)
legend("topleft",fill=c("red","blue"), legend = c("0","1"), bty="n")
```

## Extract values
### Rasters
```{r extract points from rasters}
topo.vals <- extract(topo.cp, juos, method="simple")
```

It's always a good idea to check on your data, since you can't see it. This shows us the number of values extracted from rasters. Then we can compare it with the number of points.
```{r number of points}
dim(topo.vals)
dim(coordinates(juos))
```

This shows the number of rows and columns of coordinate data, so we want it to be the same number of rows as the values above, and 2 columns wide, for x and y coordinates. Alternatively, you can just test if the number of rows are equal with the `==` operator.
```{r test if same number of points}
nrow(coordinates(juos)) == nrow(topo.vals)
```

### Vectors
Even though `crop` and `extract` are from the raster package, they have been made to work on vector data, too.
```{r extract points from vector}
soil.vals <- extract(soil.cp, juos, method="simple")
```

Let's look at the dimensions of the data frame we just created. Because the vector data represent many soil characteristics, it has created a data frame that contains many columns.
```{r dimensions of data frame extracted from soil vector}
dim(soil.vals)
```

## Combine extracted values into a data frame
```{r create data frame}
extract_juos <- data.frame(juos$coords.x1, 
                      juos$coords.x2, 
                      juos$juos, 
                      topo.vals,
                      soil.vals)
```

Now take a look at the top end of our data frame.
```{r head of data frame}
head(extract_juos)
```

The names of the columns are less than ideal. One way to give them better names is to write them into the data frame in the first place.
```{r create labeled data frame}
extract_juos <- data.frame(X=juos$coords.x1, 
                      Y=juos$coords.x2, 
                      juos=juos$juos, 
                      topo.vals,
                      soil.vals)
```

## Export data
Now export the data as a .csv file for storing and accessing.
```{r write data frame to file}
write.csv(extract_juos, "data/outdata/extract_juos.csv")
```

Take a look at the `.csv` file in MS Excel or another spreadsheet application. You'll notice that the first column contains row numbers, which we probably don't really want. So let's find out how to get rid of those. Let's use the quickest way to get help with functions (which I use very often).
```{r write.csv help}
?write.csv
```

The documentation for `write.csv` will help you to figure out that `row.names=FALSE` will allow you to get rid of that first column, so let's do that.
```{r write data frame to file without column 1}
write.csv(extract_juos, "data/outdata/extract_juos.csv", row.names=F)
```

Let's export our other files. You can export the rasters into the same file, as long as you use
a file type that supports multiple bands, like `.tif`. 
```{r write single raster}
writeRaster(topo.cp, "data/outdata/topo_cp.grd", format="raster", overwrite=T)
```

Write soils data
```{r write vector data}
writeOGR(soil.cp, "data/outdata/soil", "soil", "ESRI Shapefile", overwrite=T)
```
